\Title

### 2013-06-05 synmut.tex
streamlining discussion

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that, in the region of the V loops, a large fraction of these mutations
is deleterious. Despite their fitness cost, deleterious synonymous mutations
rise in frequency in the viral population via genetic hitchhiking due to limited
recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We show a strong correlation between the fate of a
synonymous mutation and the surrounding RNA structure, suggesting a
role of RNA hairpins around V-loops in HIV-1 evolution. We then compare our
observations to computational models and derive estimates for the effect of
synonymous mutations on viral fitness. Extending the analysis of fixation
probabilities to the nonsynonymous mutations, we show that time-dependent
selection or strong competition of escape mutations inside the same epitope are
necessary to explain the observed patterns of fixation and loss.

@@

Eventually, all SNVs will either fix or be lost. Given the SNV is at a
certain frequency $\nu$, the probability of fixation is higher for
beneficial SNVs than for neutral ones, which in turn fix more frequently
than deleterious ones. Since a neutral mutation, by definition, has no
effect on the fate of the viruses carrying it, the fixation probability 
of a neutral SNV at frequency $\nu$ is $\pfix(\nu) = \nu$; 
it goes extinct with probability $1-\nu$. For instance, if a
neutral SNV is observed in half of the sequences, it will fix with a
probability of 50\% (see inset in figure \ref{fig:aftsyn}). The fixation
probability of neutral SNVs is independent of most model assumptions
and is only affected if neutral SNVs are associated preferentially with
either high or low fitness virus.

\FIG{aft} shows the time course of the frequencies of all synonymous
SNVs (top) and nonsynonymous SNVs (bottom) observed in \env, in
\shankaregion, in a chronically HIV-1 infected patient (p10 from
\citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}). This observation seems at odds with
the assumption of neutrality.

@@

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems. We aligned the within-patient sequence samples to
the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
SHAPE reactivities to most positions in the alignment. We then calculated the
distributions of SHAPE reactivities for synonymous SNVs that fixed or were
subsequently lost (only SNVs with frequencies above 15\%). As shown in
\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
distribution, $p\approx 0.002$). In other words, SNVs that are likely to
break RNA helices are also more likely to revert and finally be lost from the
population, restoring the helix. Note that this analysis will be sensitive only
at positions where the base pairing pattern of NL4-3 agrees with that of each
patient's initial consensus sequence -- it is thus statistically conservative.
As a control, we also calculate the distribution of SHAPE reactivities for SNVs
that never reach high frequencies (green). This set is a mixture of neutral and
deleterious SNVs and, as expected, their distribution lies between those of
fixed and lost high-frequency SNVs.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the parameters in
certain bounds, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.002$, consistent with the fixation/extinction times estimated above
(see \FIG{fixp1}).

@@

The above simulations show that hitchhiking can explain the observation of
deleterious SNVs that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous SNVs at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. We
test two possible mechanisms that are biologically plausible and could explain
the transient rise of nonsynonymous SNVs: time-dependent selection and
within-epitope competition.

The former hypothesis can be formulated as follows: if the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In
support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
antibody responses to escape mutants have been reported. These responses are
delayed by a few months, roughly matching the average time needed by an escape
mutant to rise from low to high frequency. To model this type of behavior, we
assumed that antibody responses against escape SNVs arise with a rate
proportional to the frequency of the escape SNV and abolish the benefit of the
escape mutations. As expected, this type of time-dependent selection retained the
potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
diagonal as the rate of recognition of escape mutants is increased.

In the alternative hypothesis, several different escape SNVs within the same
epitope might arise almost simultaneously and start to spread. Their benefits
are not additive, because each of them is essentially sufficient to escape and
no additional benefit is gained from combining them. As a consequence, several
escape SNVs rise to high frequency rapidly, while the one with the smallest cost
in terms of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape SNVs in HIV-1
infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
Similarly, this scenario has been explicitly observed in the evolution of
resistance to 3TC, where the mutation M184V is often preceeded by M184I
\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
in the model by allowing multiple escape mutations per epitope that do
not provide additional benefit to the virus when combined. Again, we found that the potential for
hitchhiking is retained by within epitope competition but that the
fixation probability of nonsynonymous SNVs is reduced. With roughly six
mutations per epitope, the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and possibly
both important in HIV-1 evolution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus suited to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions  despite the
fact that the corresponding sites are not conservered in
cross-sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Hence selection to maintain these structures reduces the
fixation probability and hence the rate of evolution at synonymous
sites, but the molecular architecture of the structure changes over time.
As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007}. This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes finding the 
compensatory mutations that would restore base pairing in the
longitundinal data difficult, since the exact base pairing pattern is
most likely different than in the reference sequence.

A direct approach to test the function of the insulating
stems would be \textit{in vitro} replication assays on synonymous variant
viruses with unstable hairpins. This has been attempted recently, but no major
fitness effect was measured \citep{knoepfel_role_2013}. This result is
compatible with our findings that synonymous SNVs are weakly
deleterious: it would take hundreds of cell culture passages to detect fitness
effects of the order of one per mille, which is what we estimate in our
long-term dataset. The longitudinal data, however, spans many years and 
our analysis is able to quantify the subtle fitness effect of RNA structure
within single infections.

Our analysis of fixation and loss of variants also has implications for
non-synonymous SNVs observed in \env. Nonsynonymous diversity exceeds
synonymous diversity despite the overall much greater constraints on the
amino acid sequence, suggesting that the majority of high frequency SNVs
are escape mutations. If those mutations were
unconditionally beneficial, we would expect them to fix almost surely.
However, we found that many of them are lost and that they on average
show fixation behavior like neutral mutations. We suggest that this
paradoxical behavior could be due to escape
mutations that revert after they themselves are recognized by nAbs, and
the competition between different escape mutations within one epitope. 
Both mechanisms reduce the overall fixation probability and could
explain the observed pattern of fixation.

@@

While classical population genetics assumes that the dominant stochastic force
is genetic drift, i.e., non-heritable fluctuations in offspring number, our
results show that stochasticity due to linked selection is much more important.
Such fluctuations have been termed \emph{genetic draft} in
\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
large population sizes are compatible with low diversity and fast coalescence
when draft dominates over drift.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sequence data collection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Longitudinal intrapatient viral RNA sequences were collected from published
studies \citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
show substantial population structure and were discarded (see
\figurename~S\PCApat); a total of 11 patients with 4-23 time points each and
approximately 10 sequences per time point were analyzed. The time intervals
between two consecutive sequences ranged from 1 to 34 months, most of them
between 6 and 10 months.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sequence analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The sequences were translated and the resulting amino acid sequences aligned
using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
sequences separately for each patient. Within each patient, the consensus
nucleotide sequence at the first time point was used to classify alleles as
``ancestral'' or ``derived'' at all sites. Sites that include large
frequencies of gaps were excluded from the analysis to avoid artifactual
substitutions due to alignment errors. Allele frequencies at different time
points were extracted from the multiple sequence alignment.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimates of time to fixation/extinction, polymorphisms were binned by
frequency and the time to first reaching either fixation or extinction was
stored. The fixation probability was determined as the long-time limit of the
resulting curves. Mutations that reached high frequency but neither fixed nor
were lost were classified as ``floating'', with one exception: if they first
reached high frequencies within 3 years of the last time point, it was assumed
they had not had sufficient time to settle, so they were discarded.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. We used a generation time of 1
day, an outcrossing rate of $r=0.01$ per day \citep{batorsky_estimate_2011,
neher_recombination_2010}, a mutation rate of $\mu=10^{-5}$
\citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
evolution for 6000 days. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects 0.02. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with an exponentially distributed escape rate
with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
were performed and yielded comparable results.

@@

For the models with competition within epitopes, a complex epistatic fitness
landscape was designed such that each single mutant is sufficient for full
escape. In particular, each mutation had a linear effect equal to the escape,
but a negative epistatic effect of the same magnitude between each pair of sites
was included. Higher order terms compensated each other to make sure that not
only double mutants, but all k-mutants with $k \geq 1$ had the same fitness (see
supplementary materials). To model recognition of escape variants by the immune
system  catching up, the beneficial effect of an escape mutation was set
to its previous cost of -0.02 with a probability per generation proportional to
the frequency of the escape variant.

>>>>

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that, in the region of the V loops, a large fraction of these mutations
is deleterious. Despite their fitness cost, deleterious synonymous mutations
rise in frequency in the viral population via genetic hitchhiking due to limited
recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We show a strong correlation between the fate of a
synonymous mutation and the surrounding RNA structure. We then compare our
observations to computational models and obtain estimates for the effect of
synonymous mutations on viral fitness. Extending the analysis of fixation
probabilities to the nonsynonymous mutations, we show that time-dependent
selection or strong competition of escape mutations inside the same epitope are
necessary to explain the observed patterns of fixation and loss.

@@

Eventually, all SNVs will either fix or be lost, at least in small
samples. Given the SNV is at a certain frequency $\nu$, the probability
of fixation is higher for beneficial SNVs than for neutral ones, which
in turn fix more frequently than deleterious ones. Since a neutral
mutation, by definition, has no effect on the fate of the viruses
carrying it, the fixation probability of a neutral SNV at frequency
$\nu$ is $\pfix(\nu) = \nu$; it goes extinct with probability
$1-\nu$. For instance, if a neutral SNV is observed in half of the
sequences, it will fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

\FIG{aft} shows the time course of the frequencies of all synonymous
SNVs (top) and nonsynonymous SNVs (bottom) observed in the
\shankaregion{} region of \env{} in a chronically HIV-1 infected patient (p10 from
\citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}). This observation seems at odds with
the assumption of neutrality.

@@

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases 
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems. We aligned the within-patient sequence samples to
the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
SHAPE reactivities to most positions in the alignment. We then calculated the
distributions of SHAPE reactivities for synonymous SNVs that fixed or were
subsequently lost (only SNVs with frequencies above 15\%). As shown in
\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
distribution, $p\approx 0.002$). In other words, SNVs that are likely to
break RNA helices are also more likely to revert and finally be lost from the
population, restoring the helix. Note that this analysis will be sensitive only
at positions where the base pairing pattern of NL4-3 agrees with that of each
patient's initial consensus sequence -- it is thus statistically conservative.
As a control, we also calculate the distribution of SHAPE reactivities for SNVs
that never reach high frequencies (green). This set is a mixture of neutral and
deleterious SNVs and, as expected, their distribution lies between those of
fixed and lost high-frequency SNVs.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the
parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.002$, consistent with the fixation/extinction times estimated above
(see \FIG{fixp1}).

@@

The above simulations show that hitchhiking can explain the observation of
deleterious SNVs that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous SNVs at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

If the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In
support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
antibody responses to escape mutants have been reported. These responses are
delayed by a few months, roughly matching the average time needed by an escape
mutant to rise from low to high frequency. To model this type of behavior, we
assumed that antibody responses against escape SNVs arise with a rate
proportional to the frequency of the escape SNV and abolish the benefit of the
escape mutations. As expected, this type of time-dependent selection retained the
potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
diagonal as the rate of recognition of escape mutants is increased.

Several different escape mutations that arise within the 
same epitope have similar effects. Their benefits
are not additive, because each of them is essentially sufficient to escape and
no additional benefit is gained from combining them. As a consequence, several
escape SNVs rise to high frequency rapidly, while the one with the smallest cost
in terms of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape SNVs in HIV-1
infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
Similarly, this scenario has been explicitly observed in the evolution of
resistance to 3TC, where the mutation M184V is often preceeded by M184I
\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
in the model by allowing multiple escape mutations per epitope that do
not provide additional benefit to the virus when combined. Again, we found that the potential for
hitchhiking is retained by within epitope competition but that the
fixation probability of nonsynonymous SNVs is reduced. With roughly six
mutations per epitope, the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and possibly
both important in HIV-1 evolution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonynous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years and our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus suited to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions  despite the
fact that the corresponding sites are not conservered in
cross-sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of the structure
changes over time, selection seems to maintain these structures, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007}. This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes finding the 
compensatory mutations that would restore base pairing in the
longitundinal data difficult, since the exact base pairing pattern is
most likely different than in the reference sequence.

Selection against the majority of synonymous substitutions is probably
true across the genome, but we only observe deleterious synonymous SNVs
at high frequency in \shankaregion{} of \env, where they hitch-hike to
high frequencies on nAb escape mutations. Surprisingly, that non-synonymous
mutations display a fixation pattern as if they were neutral;
comp.~\FIG{pfix}. However, nonsynonymous diversity exceeds
synonymous diversity despite the overall much greater constraints on the
amino acid sequence, suggesting that the majority of high frequency SNVs
are escape mutations despite the fact that are often lost again. We suggest that this
paradoxical behavior could be due to escape
mutations that revert after they themselves are recognized by nAbs, and
the competition between different escape mutations within one epitope. 
Both mechanisms reduce the overall fixation probability and could
explain the observed pattern of fixation.

@@

While classical population genetics assumes that the dominant stochastic force
is genetic drift, i.e., non-heritable fluctuations in offspring number, our
results show that stochasticity due to linked selection is much more important.
Such fluctuations have been termed \emph{genetic draft} in
\citet{gillespie_genetic_2000} and their effect in facultatively sexual population
such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
large population sizes are compatible with low diversity and fast coalescence
when draft dominates over drift.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sequence data collection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Longitudinal intrapatient viral RNA sequences were collected from published
studies \citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. Some patients
show substantial population structure and were excluded (see
\figurename~S\PCApat); a total of 11 patients with 4-23 time points each and
approximately 10 sequences per time point were analyzed. The time intervals
between two consecutive sequences ranged from 1 to 34 months, most of them
between 6 and 10 months.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sequence analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The sequences were translated and the resulting amino acid sequences aligned
using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
sequence separately for each patient. Within each patient, the consensus
nucleotide sequence at the first time point was used to classify alleles as
``ancestral'' or ``derived'' at all sites. Sites that include large
frequencies of gaps were excluded from the analysis to avoid artifactual
substitutions due to alignment errors. Allele frequencies at different time
points were extracted from the multiple sequence alignment.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimates of time to fixation/extinction, SNVs were binned by
frequency and the time to first reaching either fixation or extinction was
stored. The fixation probability was determined as the long-time limit of the
resulting curves. Mutations that reached high frequency but neither fixed nor
were lost were classified as ``floating'', with one exception: if they first
reached high frequencies within 3 years of the last time point, it was assumed
they had not had sufficient time to settle, so they were discarded.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. We used a generation time of 1
day, an outcrossing rate of $r=0.01$ per day \citep{batorsky_estimate_2011,
neher_recombination_2010}, a mutation rate of $\mu=10^{-5}$
\citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
evolution for 6000 days. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects 0.02. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
were performed and yielded comparable results.

@@

For the models with competition within epitopes, a complex epistatic fitness
landscape was designed such that each single mutant is sufficient for full
escape. Specifically, each mutation has an additive effect equal to the
escape rate, but interacts with all other escape mutations in the same
epitope with a negative effect of the same magnitude. 
Higher order terms were included to make sure that not
only double mutants, but all k-mutants with $k \geq 1$ had the same fitness (see
supplementary materials). To model recognition of escape variants by the
evolving immune system, the beneficial effect of an escape mutation was set
to its previous cost of -0.02 with a probability per generation proportional to
the frequency of the escape variant.
